---
title: "CD4 Cell Count - Longitudinal Data Analysis   "
output: html_notebook
---

```{r}
library(ggplot2)
library(dplyr)
library(splines)
library(nlme)
library(GGally)
```

## Introduction

The average interval from infection by the human immune-deficiency virus (HIV) to AIDS is estimated to be approximately 11 years. major difficulty in studying progression of HIV infection is that an unreasonably long follow-up is necessary to observe a large cohort over the entire incubation period from infection to disease. An alternative to monitoring the presence or absence of AIDS is to follow a surrogate marker of disease progression for shorter intervals.

HIV destroys T-lymphocytes called CD4 cells, which play a vital role in immune function. Disease progression can be assessed by measuring the number or percent of CD4 cells, which on average decrease throughout the disease incubation period.

The project uses the CD4 dataset, the response variable is CD4+ cell count. The dataset consists of longitudinally collected observations on 369 subjects, resulting in a total of
2376 observations of CD4 cell counts denoted CD4 in the dataset. Other variables collected are:

1. Time: as the time (in years) since seroconversion, where a negative time denotes actual time
before seroconversion.
2. Age: age at seroconversion (a baseline measurement), centred at 30 years of age, so that negative
ages denote years younger than 30.
3. Packs: the number of packets of cigarettes smoked per day at time of measurement.
4. Drugs: a binary variable taking the values 1 or 0 to denote if the respondent takes recreational
drugs or not respectively, measured at each time point.
5. Sex: number of sexual partners reported at each time point. Looks to have been centred somehow
and truncated at ±5.
6. Cesd: an index of depression measured at each time point, with time trends removed. Higher
scores indicate greater depressive symptoms.

## Exploratory Data Analysis

```{R}
# Load data
data <- read.table("./cd4data.txt", header = TRUE)
data$CD4sqrt <- sqrt(data$CD4)
```

```{r}
ggpairs(data = data[,c(1, 3, 4, 5, 6,7 ,9)])
```

In the first step of exploratory data anslysis of CD4 data, we plot a pairs plot of the dataset. From the plot it can be seen that the square root of CD4 cell count is roughly normally
distributed. Therefore, square root of CD4 cell count will be used as response variable in
further analysis. The CD4 cell count is negatively correlated with time sand slightly positively
correlated with Packs, Drugs and Sex variables.

```{r}
p1 <- ggplot(data = data, aes(x = Time, y = CD4sqrt)) + 
  geom_point() + 
  geom_smooth(method = 'loess', se=FALSE) + 
  ggtitle("sqrtCD4 vs Time")
print(p1)
```

The CD4+ cell number are constant until the time of seroconversion i.e. Time = 0,
after that the cell number appear to be decreasing. Overall, there seems to be a decreasing
trend in the CD4 cell count with time.

```{r}
# plot smoothed curve of cell count wrt drug use
df1 <- data %>% filter(Drugs==0) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))
df2 <- data %>% filter(Drugs==1) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))

p1 <- ggplot() + geom_smooth(data = df1, aes(x = Time, y = mean_cd4), method=
                               "loess", se=FALSE) + 
  geom_smooth(data = df2, aes(x = Time, y = mean_cd4), method="loess", se=FALSE, color="red") + ggtitle("Mean sqrtCD4 vs Time (Drug use)") + ylab("Mean sqrtCD4")
plot(p1)
```

```{r}
df1 <- data %>% filter(Age <= quantile(data$Age, 0.25)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))
df2 <- data %>% filter(Age >= quantile(data$Age, 0.75)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))

p2 <- ggplot() + geom_smooth(data = df1, aes(x = Time, y = mean_cd4), method=
                               "loess", se=FALSE) + 
  geom_smooth(data = df2, aes(x = Time, y = mean_cd4), method="loess", se=FALSE, color="red") + ggtitle("Mean sqrtCD4 vs Time (Age)") + ylab("Mean sqrtCD4")
plot(p2)
```

```{r}
df1 <- data %>% filter(Cesd <= quantile(data$Cesd, 0.25)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))
df2 <- data %>% filter(Cesd >= quantile(data$Cesd, 0.75)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))

p3 <- ggplot() + geom_smooth(data = df1, aes(x = Time, y = mean_cd4), method=
                               "loess", se=FALSE) + 
  geom_smooth(data = df2, aes(x = Time, y = mean_cd4), method="loess", se=FALSE, color="red") + ggtitle("Mean sqrtCD4 vs Time (CESD)") + ylab("Mean sqrtCD4")
plot(p3)
```

```{r}
df1 <- data %>% filter(Packs <= quantile(data$Packs, 0.25)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))
df2 <- data %>% filter(Packs >= quantile(data$Packs, 0.75)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))

p4 <- ggplot() + geom_smooth(data = df1, aes(x = Time, y = mean_cd4), method=
                               "loess", se=FALSE) + 
  geom_smooth(data = df2, aes(x = Time, y = mean_cd4), method="loess", se=FALSE, color="red") + ggtitle("Mean sqrtCD4 vs Time (Smoking)") + ylab("Mean sqrtCD4")
plot(p4)
```

There seems to be a clear gap in the mean responses of CD4 cell count when
stratified by drug use and smoking. Age and Cesd index seems to have no effect on the CD4
cell count. The effects of these variables will be studied in the preliminary models discussed
in the next section.
  