---
title: "CD4 Cell Count - Longitudinal Data Analysis   "
output: html_notebook
---

```{r}
library(ggplot2)
library(dplyr)
library(splines)
library(nlme)
library(GGally)
```

## Introduction

The average interval from infection by the human immune-deficiency virus (HIV) to AIDS is estimated to be approximately 11 years. major difficulty in studying progression of HIV infection is that an unreasonably long follow-up is necessary to observe a large cohort over the entire incubation period from infection to disease. An alternative to monitoring the presence or absence of AIDS is to follow a surrogate marker of disease progression for shorter intervals.

HIV destroys T-lymphocytes called CD4 cells, which play a vital role in immune function. Disease progression can be assessed by measuring the number or percent of CD4 cells, which on average decrease throughout the disease incubation period.

The project uses the CD4 dataset, the response variable is CD4+ cell count. The dataset consists of longitudinally collected observations on 369 subjects, resulting in a total of
2376 observations of CD4 cell counts denoted CD4 in the dataset. Other variables collected are:

1. Time: as the time (in years) since seroconversion, where a negative time denotes actual time
before seroconversion.
2. Age: age at seroconversion (a baseline measurement), centred at 30 years of age, so that negative
ages denote years younger than 30.
3. Packs: the number of packets of cigarettes smoked per day at time of measurement.
4. Drugs: a binary variable taking the values 1 or 0 to denote if the respondent takes recreational
drugs or not respectively, measured at each time point.
5. Sex: number of sexual partners reported at each time point. Looks to have been centred somehow
and truncated at ±5.
6. Cesd: an index of depression measured at each time point, with time trends removed. Higher
scores indicate greater depressive symptoms.

## Exploratory Data Analysis

```{R}
# Load data
data <- read.table("./cd4data.txt", header = TRUE)
data$CD4sqrt <- sqrt(data$CD4)
```

```{r}
ggpairs(data = data[,c(1, 3, 4, 5, 6,7 ,9)])
```

In the first step of exploratory data anslysis of CD4 data, we plot a pairs plot of the dataset. From the plot it can be seen that the square root of CD4 cell count is roughly normally
distributed. Therefore, square root of CD4 cell count will be used as response variable in
further analysis. The CD4 cell count is negatively correlated with time sand slightly positively
correlated with Packs, Drugs and Sex variables.

```{r}
p1 <- ggplot(data = data, aes(x = Time, y = CD4sqrt)) + 
  geom_point() + 
  geom_smooth(method = 'loess', se=FALSE) + 
  ggtitle("sqrtCD4 vs Time")
print(p1)
```

The CD4+ cell number are constant until the time of seroconversion i.e. Time = 0,
after that the cell number appear to be decreasing. Overall, there seems to be a decreasing
trend in the CD4 cell count with time.

```{r}
# plot smoothed curve of cell count wrt drug use
df1 <- data %>% filter(Drugs==0) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))
df2 <- data %>% filter(Drugs==1) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))

p1 <- ggplot() + geom_smooth(data = df1, aes(x = Time, y = mean_cd4), method=
                               "loess", se=FALSE) + 
  geom_smooth(data = df2, aes(x = Time, y = mean_cd4), method="loess", se=FALSE, color="red") + ggtitle("Mean sqrtCD4 vs Time (Drug use)") + ylab("Mean sqrtCD4")
plot(p1)
```

```{r}
df1 <- data %>% filter(Age <= quantile(data$Age, 0.25)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))
df2 <- data %>% filter(Age >= quantile(data$Age, 0.75)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))

p2 <- ggplot() + geom_smooth(data = df1, aes(x = Time, y = mean_cd4), method=
                               "loess", se=FALSE) + 
  geom_smooth(data = df2, aes(x = Time, y = mean_cd4), method="loess", se=FALSE, color="red") + ggtitle("Mean sqrtCD4 vs Time (Age)") + ylab("Mean sqrtCD4")
plot(p2)
```

```{r}
df1 <- data %>% filter(Cesd <= quantile(data$Cesd, 0.25)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))
df2 <- data %>% filter(Cesd >= quantile(data$Cesd, 0.75)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))

p3 <- ggplot() + geom_smooth(data = df1, aes(x = Time, y = mean_cd4), method=
                               "loess", se=FALSE) + 
  geom_smooth(data = df2, aes(x = Time, y = mean_cd4), method="loess", se=FALSE, color="red") + ggtitle("Mean sqrtCD4 vs Time (CESD)") + ylab("Mean sqrtCD4")
plot(p3)
```

```{r}
df1 <- data %>% filter(Packs <= quantile(data$Packs, 0.25)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))
df2 <- data %>% filter(Packs >= quantile(data$Packs, 0.75)) %>% group_by(Time) %>% summarise(mean_cd4 = mean(CD4sqrt))

p4 <- ggplot() + geom_smooth(data = df1, aes(x = Time, y = mean_cd4), method=
                               "loess", se=FALSE) + 
  geom_smooth(data = df2, aes(x = Time, y = mean_cd4), method="loess", se=FALSE, color="red") + ggtitle("Mean sqrtCD4 vs Time (Smoking)") + ylab("Mean sqrtCD4")
plot(p4)
```

There seems to be a clear gap in the mean responses of CD4 cell count when
stratified by drug use and smoking. Age and Cesd index seems to have no effect on the CD4
cell count. The effects of these variables will be studied in the preliminary models discussed
in the next section.

# Model Formulation

To start with the model selection and analysis, we use the full model with seven knots as
given in Zeger and Diggle (1994). According to the fit of the full model, the Age and Sex
variables are not significant at 5% level of significance. Therefore, in further models, these
variables are excluded. The full model is further simplified to decrease the number of knots
and one model with no knots. The following preliminary models are considered:
1. Full model with all the variables and seven knots at equal intervals
2. Model with variables: smoking, Cest index, Drug and time, with two knots at T=0,2
3. Model with variables: Smoking, Cest index, Drug and time
4. Model with variables: Smoking, Cest index, Drug and time with one knot at T=0

```{r}
# Full model with 7 knots
Time.range <- range(data$Time)
n.knots <- 7
knot.spacing <- diff(Time.range)/(n.knots+1)
Time.knots <- seq(Time.range[1], Time.range[2], knot.spacing)[-c(1,(n.knots+2))]
model1 <- lm(CD4sqrt ~ ns(Time,knots = Time.knots) + 
               Age + Packs + Drugs + Cesd + Sex, data = data)
summary(model1)
```

```{r}
# Model with two knots at Time = 0 and Time = 2 and excluding Age and Sex
model2 <- lm(CD4sqrt ~ ns(Time,knots = c(0, 2)) + 
               Packs + Drugs + Cesd, data = data)
summary(model2)
```

```{r}
# Model with four predictor variables
model3 <- lm(CD4sqrt ~ Time + Packs + Drugs + Cesd, data = data)
summary(model3)
```

```{r}
# Piecewise linear model
data$Timesp <- data$Time*(data$Time >= 0)
model4 <- lm(CD4sqrt ~ Time + Timesp + Packs + Drugs + Cesd, data = data)
summary(model4)
```

```{r}
# Comparing the models
AIC(model1, model2, model3, model4)
```

The full model with seven knots is slightly better performing model among all the models considered. The model with four variables is better performing than the model with one knot at Time=0. We will choose model with four variables: Time, Smoking, Drugs and Cesd for further analysis as there is not much difference in all the models considered. 

We can see that all the variables considered in this model are significant at 5% level of significance. This model can also be written as:

$$
sqrtCD4 = 26.29 − 1.621 ×    Time + 0.989 × Packs + 1.059 × Drugs − 0.034 × Cesd
$$

The coefficients for Time and Cesd are negative, which means that CD4 cell count is
decreasing over time and with increasing Cesd score. The coefficients for Packs and Drugs is
positive, which means that the cell count is increasing with increased smoking and drug use.

```{r warning=FALSE}
# Model 3 residuals
data$resid <- resid(model3)
data$yr <- round(data$Time)
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y, use = "pairwise.complete.obs"))
  txt <- format(c(r, 0.123456789), digits=digits)[1]
  txt <- paste(prefix, txt, sep="")
  if(missing(cex.cor)) cex <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex * r)
}
data.wide <- reshape(data[,c("ID", "resid", "yr")],
  direction = "wide", v.names = "resid",
  timevar = "yr", idvar = "ID")
pairs(data.wide[,c(5, 2, 3, 6:8)], upper.panel = panel.cor)
```

The correlation is decreasing for the observations that are further apart in time.

```{r}
# Variogram
vijk <- by(data, data$ID, function(df) {
  v <- outer(df$resid, df$resid,
  function(x, y) 0.5*(x-y)^2)
  v[lower.tri(v)]
})

uijk <- by(data, data$ID, function(df) {
  u <- outer(df$Time, df$Time,
  function(x, y) abs(x - y))
  u[lower.tri(u)]
})

uijk <- unlist(uijk)
vijk <- unlist(vijk)

vu.lowess <- lowess(uijk, vijk)
sigma2 <- var(data$resid)

plot(uijk, vijk, col = "gray50", pch = 18, cex = 0.4,
  xlim = c(0, 6), ylim = c(0, 150),
  xlab = "Lag", ylab = "Half squared differences")
lines(vu.lowess, col = "red", lwd = 2)
abline(h = sigma2, lty = 2)
```
The variogram is increasing with increase in lag, which also shows that the correlation is decreasing when observations are further apart in time.

Further in the analysis, we fit model with different covariance structures for patients and
evaluate the fit. The following models are considered:
1. Model with compound symmetry covariance structure.
2. Model with unequal variances for each patient.
3. Model with exponential covariance structure.
4. Model with Gaussian covariance structure.

```{r}
data <- data %>% group_by(ID) %>%
            mutate(obsnum = seq_along(ID)) %>%
            ungroup()
```   

```{r}
# Compound Symmetry
modcov.cc <- gls(CD4sqrt ~ Time + Packs + Drugs + Cesd,
                    correlation = corCompSymm(form = ~ obsnum | ID),
                    weights = varIdent(form = ~ 1 | obsnum),
                    data = data)
summary(modcov.cc)
```


```{r}
# Unequal variances
modcov.uv <- gls(CD4sqrt ~ Time + Packs + Drugs + Cesd,
                    weights = varIdent(form = ~ 1 | obsnum),
                    data = data)
summary(modcov.uv)
```

```{r}
# Exponential covariance structure
modcov.exp <- gls(CD4sqrt ~ Time + Packs + Drugs + Cesd,
                    correlation = corExp(form = ~ obsnum | ID),
                    data = data)
summary(modcov.exp)
```

```{r}
# Gaussian covariance structure
modcov.gaus <- gls(CD4sqrt ~ Time + Packs + Drugs + Cesd,
                    correlation = corGaus(form = ~ obsnum | ID),
                    data = data)
summary(modcov.gaus)
```

```{r}
# comparing models
AIC(model1, model3, modcov.cc, modcov.uv, modcov.exp, modcov.gaus)
```

From the table, it can be seen that the model with exponential covariance structure fits the data adequately. The value of estimated parameter θ in the exponential model is 2.077. Also, comparing with the preliminary models, these models perform better.

Now, we will consider adding random effects to the fixed effects already modeled previously.
The random effect models include random effect for time variable. The following
models are considered:
1. Random effect for Time with no correlation structure.
2. Random effect for Time with exponential correlation structure.
3. Random effect for Time with Gaussian correlation structure.

```{r}
# Random effect model
mod.rand1 <- lme(CD4sqrt ~ Time + Packs + Drugs + Cesd, 
                 random  = ~ Time | ID,
                 data = data)

summary(mod.rand1)
```

```{r}
# Random effect model with Gaussian correlation structure
mod.gau.rand <- lme(CD4sqrt ~ Time + Packs + Drugs + Cesd, 
                    random= ~ Time | ID, corr=corGaus(, form= ~ Time | ID,nugget=TRUE,value=c(2,.1)),
data=data)
summary(mod.gau.rand)
```

```{r}
# Random effect model with exponential correlation structure
mod.exp.rand <- lme(CD4sqrt ~ Time + Packs + Drugs + Cesd, 
                    random= ~ Time | ID, corr=corExp(, form= ~ Time | ID,nugget=TRUE,value=c(2,.1)),
data=data)
summary(mod.exp.rand)
```

```{r}
# Comparing models  
AIC(mod.exp.rand, mod.gau.rand, mod.rand1)
```

```{r}
# Final model
mod <- lme(CD4sqrt ~ Time + Packs + Drugs + Cesd, 
                    random= ~ Time | ID, corr=corGaus(, form= ~ Time | ID,nugget=TRUE,value=c(2,.1)),
data=data)
summary(mod)
```   

Comparing the AIC values for all the models considered so far, we can say that the
random effect model with Gaussian correlation structure is a better model. The Gaussian correlation structure model with random effects can also be written as:

$$
sqrtCD4 = β_1 + β_2 \times Time + β_3 × Packs + β_4 × Drugs + β_5 × Cesd + b_1 + b_2 \times Time
$$

The variance of parameter b1 is 3.707 and the variance of parameter b2 is 1.144. The
correlation between the two random effect variables is 0.103.

### Application to Individual Trajectories

A patient was selected from the CD4 data, with measurement spread roughly evenly over
Time = 0.

```{R}
# find patient
selected <- data %>% 
  filter((Time <= -1.5 & obsnum >= 3) | (Time >= 1.5 & obsnum >= 6)) 

IDs <- unique(selected$ID)
sel.data <- data %>% filter(ID %in% IDs)

ggplot(data = sel.data[sel.data$ID==10569,], aes(x=Time, y = CD4sqrt)) + 
  geom_line() + 
  ggtitle("CD4sqrt vs Time (Patient ID 10569)")
```

```{R}
# Covariance matrix
covmat <- getVarCov(mod, individual = "10569", type = 'marginal')
cormat <- round(cov2cor(covmat$"10569"), 3)
cormat
```
From the correlation matrix for the selected patient, we can see that the
correlation is decreasing with time of observation.

Next, we select three patients using the same criteria. We use the Best Linear Unbiased
Predictor (BLUP) and the chosen model to predict the individual trajectories.

```{r}
# BLUP
patient1 <- subset(data, ID == 10569)
patient2 <- subset(data, ID == 11172)
patient3 <- subset(data, ID == 20363)

plot(CD4sqrt~Time,data=patient1,pch=4, col="red",xlim=c(-3,3),xlab="Time",ylab="sqrtCD4")
points(CD4sqrt~Time,data=patient2,pch=3, col="blue")
points(CD4sqrt~Time,data=patient3,pch=3, col="green")

t3 <- seq(-3,3,0.01)
t2 <- seq(-3,3,0.01)
t1 <- seq(-3,3,0.01)
t0 <- seq(-3,3,0.01)
incpt<-rep(1,length(t1))
X<-cbind(incpt,t3,t2,t1,t0)
incpt<- cbind(incpt, t0)

pmean<-X%*%mod$coefficients$fixed

lines(t1,pmean,lty=1,lwd=3)
l1<-incpt%*%mod$coefficients$random$ID["10569",]+pmean
lines(t1,l1,col="red",lty=1,lwd=2)

l2<-incpt%*%mod$coefficients$random$ID["11172",]+pmean
lines(t1,l2,col="blue",lty=1,lwd=2)

l3<-incpt%*%mod$coefficients$random$ID["20363",]+pmean
lines(t1,l3,col="green",lty=1,lwd=2)
```

The observed level of CD4 cell counts are also shown on the plot. As the random effect model has random intercept as well as random slope for time, which is shown in the separate trajectories.

In this project, we used exploratory data analysis to visualize several predictor variables with
respect to the response variable, CD4 cell count. In the preliminary models, the spline model
with six knots was slightly better than the other simpler models. A simple model with linear
variables with no knots was chosen for further analysis. Several different covariance structure were fitted and compared using AIC. Model with exponential covariance structure was found
to be better.

The models including random effects and different correlation structures were analysed
next. A model with random effect on time and Gaussian correlation structure was found to
be better than all the models considered previously.